Tree biodiversity and ecosystem function relationship across southern African woodlands

### 2020-09-17 manuscript/include/clust_summ.tex
Anna's proofread

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Wed, Jul 15, 2020 - 09:38:19
\begin{table}[!htbp] \centering 
  \caption{Description of the biogeographical clusters to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest mean plot level proportional AGB within each cluster. Numeric values of species richness, stems ha\textsuperscript{-1} and AGB represent medians and interquartile ranges (75th percentile - 25th percentile).} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
{Cluster} & {Dominant species} & {Indicator species} & {N plots} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
\hline \\[-1.8ex] 
Core miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Brachystegia spiciformis} \\\textit{Julbernardia paniculata} \\\textit{Brachystegia boehmii} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Parinari curatellifolia} \\\textit{Uapaca kirkiana} \\\textit{Brachystegia spiciformis} \end{tabular} & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\ 
\hline
ex-Acacia & \begin{tabular}[c]{@{}c@{}c@{}} Spirostachys africana \\\textit{Senegalia burkei} \\\textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\\textit{Vachellia nilotica} \\\textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\ 
\hline
Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\\textit{Androstachys johnsonii} \\\textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{\textit{Colophospermum mopane}} \\\textit{Psuedolachnostylis maprouneifolia} \\\textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\ 
\hline
Sparse miombo / Baikiaea & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\\textit{Burkea africana} \\\textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\\textit{Baikiaea plurijuga} \\\textit{Pterocarpus angolensis} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\ 
\hline
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

>>>>

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Wed, Jul 15, 2020 - 09:38:19
\begin{table}[!htbp] \centering 
  \caption{Description of the biogeographical clusters to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest mean plot level proportional AGB within each cluster. Numeric values of species richness, stems ha\textsuperscript{-1} and AGB represent medians and interquartile ranges (75th percentile - 25th percentile).} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
{Cluster} & {Dominant species} & {Indicator species} & {N plots} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
\hline \\[-1.8ex] 
Core miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Brachystegia spiciformis} \\\textit{Julbernardia paniculata} \\\textit{Brachystegia boehmii} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Parinari curatellifolia} \\\textit{Uapaca kirkiana} \\\textit{Brachystegia spiciformis} \end{tabular} & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\ 
\hline
ex-Acacia & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\\textit{Senegalia burkei} \\\textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\\textit{Vachellia nilotica} \\\textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\ 
\hline
Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\\textit{Androstachys johnsonii} \\\textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{\textit{Colophospermum mopane}} \\\textit{Psuedolachnostylis maprouneifolia} \\\textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\ 
\hline
Sparse miombo / Baikiaea & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\\textit{Burkea africana} \\\textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\\textit{Baikiaea plurijuga} \\\textit{Pterocarpus angolensis} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\ 
\hline
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2020-09-17 manuscript/regional_befr_manuscript.tex
Anna's proofread

Authors: John L. Godlee\textsuperscript{1}, Casey M. Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J. Bowers\textsuperscript{1}, Joao M. B. Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P. G. M. Cromsigt\textsuperscript{6,7,8}, Dave J. Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M. Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M. McNicol\textsuperscript{1}, Edward T. A. Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G. Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Dexter, K. G.\textsuperscript{1,2}

@@

% Situation
The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealt{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen
Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.

@@

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, without an effect of species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

% Re-cap
In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots. Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}). Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level. The strongest effect on AGB was that of stem density. When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.

@@

We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}). At low stem densities competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}. At very high stem density, there is also an increase in the effect of species diversity on structural diversity. This could be because at high stem density, the adaptation of different species to growth form become important. At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species ara able to exist in the understory, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.

@@

Core miombo and sparse miombo / Baikiaea woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types. Compared to ex-Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness. ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp. in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees. We postulate that the slight negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in ex-Acacia and Mopane woodlands, where the top level of the woodland canopy is dominated often by a single species.

Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation and suppressing the few species which tend to create high biomass, such as \textit{Julbernadia} and \textit{Brachystegia} spp.. In the species poor Mopane and ex-Acacia woodlands however, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.

>>>>

Authors: John L. Godlee\textsuperscript{1}, Casey M. Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J. Bowers\textsuperscript{1}, Joao M. B. Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P. G. M. Cromsigt\textsuperscript{6,7,8}, Dave J. Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M. Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M. McNicol\textsuperscript{1}, Edward T. A. Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G. Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Kyle G. Dexter\textsuperscript{1,23}

@@

% Situation
The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen
Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.

@@

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

% Re-cap
In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots. Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}). Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level. The strongest effect on AGB was that of stem density. When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.

@@

We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}). At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}. At very high stem density, there is also an increase in the effect of species diversity on structural diversity. This could be because at high stem density, the adaptation of different species to growth form become important. At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.

@@

Core miombo and sparse miombo / Baikiaea woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types. Compared to ex-Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness. Ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp. in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees. We postulate that the slight negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in ex-Acacia and Mopane woodlands, where the top level of the woodland canopy is dominated often by a single species.

Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation and suppressing the few species which tend to create high biomass, such as \textit{Julbernadia} and \textit{Brachystegia} spp.. In the species-poor Mopane and ex-Acacia woodlands however, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
